### 2017-05-06 2017mergingmerges/section/01-Introduction.tex
Edits to S2, get state diag and fused result on same page

The @group@ operator detects groups of consecutive identical elements and emits a single representative, while @merge@ combines two sorted streams so that the output remains sorted. This example has a few interesting properties. Firstly, the data-access pattern of @merge@ is \emph{value-dependent}, meaning that the order in which @merge@ pulls values from @sIn1@ and @sIn2@ depends on the values themselves: at each step, @merge@ must compare the values from both streams, and choose the stream with the smaller value to pull from.
Secondly, although @sIn1@ occurs twice in the program, at runtime we only want to handle the elements of each stream once. To achieve this, the compiled program must coordinate between the two uses of @sIn1@, so that a new value is read only when both the @group@ and @merge@ operators are ready to receive it. Finally, as the stream length is unbounded, we cannot buffer an arbitrary number of elements read from either stream, or risk running out of local storage space.

@@

Instead, for a single threaded program, we want to perform \emph{stream fusion}, which takes the implied dataflow network and produces a simple sequential loop that does not require extra process-control abstractions or unbounded buffering. Sadly, existing stream fusion transformations cannot handle our example.

>>>>

The @group@ operator detects groups of consecutive identical elements and emits a single representative, while @merge@ combines two sorted streams so that the output remains sorted. This example has a few interesting properties. Firstly, the data-access pattern of @merge@ is \emph{value-dependent}, meaning that the order in which @merge@ pulls values from @sIn1@ and @sIn2@ depends on the values themselves: at each step, @merge@ must compare the values from both streams, and choose the stream with the smaller value to pull from.

Secondly, although @sIn1@ occurs twice in the program, at runtime we only want to handle the elements of each stream once. To achieve this, the compiled program must coordinate between the two uses of @sIn1@, so that a new value is read only when both the @group@ and @merge@ operators are ready to receive it. Finally, as the stream length is unbounded, we cannot buffer an arbitrary number of elements read from either stream, or risk running out of local storage space.

@@

Instead, we would prefer to use \emph{stream fusion}, which is a program transformation that takes the implied dataflow network and produces a simple sequential loop that does not require extra process-control abstractions or unbounded buffering. Sadly, existing stream fusion transformations cannot handle our example.

### 2017-05-06 2017mergingmerges/section/02-Processes.tex
Edits to S2, get state diag and fused result on same page

% -----------------------------------------------------------------------------
\section{Processes, Machines, Combinators and Operators}
\label{s:Processes}

@@

The definition of the @group@ combinator which detects groups of successive identical elements in the input stream is given in Fig.~\ref{fig:Process:Group}. The process emits the first value pulled from the stream and every value that is different from the last one that was pulled. For example, when executed on the input stream @[1,2,2,3]@, the process will produce the output @[1,2,3]@. We include the concrete representation and a diagram of the process viewed as a state machine.

@@

The body of the combinator is a record that defines the process. The @ins@ field of the record defines the set of input streams and the @outs@ field the set of output streams. The @heap@ field gives the initial values of each of the local variables. The @instrs@ field contains a set of labeled instructions that define the program, while the @label@ field gives the label of the initial instruction. Note that in this form the output stream is defined via a parameter, rather than being the result of the combinator, as in the source representation of @uniquesUnion@ from \S\ref{s:Introduction}.

@@

% AR: Clumsy, but added "at L2 executes:" to force @(push...)@ onto its own line. Otherwise it goes off the right.
When the predicate is true, the instruction at label @A2@ executes:
@(push sOut1 v A3 [ l = v, f = F ])@
pushes the value @v@ to the output stream @sOut1@ and proceeds to the instruction at label @A3@, once the variable @l@ is set to @v@ and @f@ to @F@ (False).

Finally, the instruction @(drop sIn1 A0 [])@ signals that the current element that was pulled from stream @sIn1@ is no longer required, and goes back to the first the instruction at @A0@. This @drop@ instruction is used to coordinate concurrent processes when performing fusion. The next element of a stream may only be pulled after all consumers of that stream have pulled and then and dropped the current element.

@@

The definition of the @merge@ combinator, which merges two input streams, is given in Fig.~\ref{fig:Process:Merge}. The combinator binds the two input streams to @sIn1@ and @sIn2@, while the output stream is @sOut2@. The two heap variables @x1@ and @x2@ are used to store the last values read from each input stream. The process starts by pulling from each of the input streams. It then compares the two pulled values, and pushes the smaller of the values to the output stream. The process then drops the stream which yielded the the smaller value, then pulls from the same stream so that it can perform the comparison again.

As the @merge@ process merges infinite streams, if we execute it with a finite input prefix, it will arrive at an intermediate state that may not yet have pushed all available output. For example, if we execute the process with the input streams @[1,4]@ and @[2,3,100]@ then the values @[1,2,3,4]@ will be pushed to the output. After pushing the last value @4@, the process will block at instruction @E2@, waiting for the next value to become available from @sIn2@. We discuss how to handle finite streams later in ~\S\ref{s:Finite}.

@@

Our fusion algorithm takes two processes and produces a new one that computes the output of both. For example, suppose we need a single process that produces the output of the first two lines of our @uniquesUnion@ example back in \S\ref{s:Introduction}. The result will be a process that computes the result of both @group@ and @merge@ as if they were executed concurrently, where the first input stream of the @merge@ process is the same as the input stream of the @group@ process. For ease of reference, when describing the fusion algorithm we will instantiate the parameters of each combinator with arguments of the same names.

@@

The algorithm proceeds by considering pairs of states: one in each of the source process state machines to be fused. Both the @group@ machine and the @merge@ machine pull from the same stream as their initial instruction, so we have the situation shown in Fig.~\ref{fig:Fusion:Pulls}. The @group@ machine needs to transition from label @A0@ to label @A1@, and the @merge@ machine from @B0@ to @B1@. In the result machine we produce three new instructions that transition between four joint result states, @F0@ to @F3@.
Each of the joint result states represents a combination of two source states, one from each of the source machines. For example, the first result state @F0@ represents a combination of the @group@ machine being in its initial state @A0@ and the @merge@ machine being in its own initial state @B0@.

We also associate each of the joint result states with whether or not each source machine has already pulled a value from each of its input streams. For the @F0@ case shown in Fig.~\ref{fig:Fusion:Pulls} we have ((A0, \{sIn1 = none\}), (B0, \{sIn1 = none, sIn2 = none\})). The result state @F0@ represents a combination of the two source states @A0@ and @B0@. As both @A0@ and @B0@ are the initial states of their respective machines, those machines have not yet pulled any values from their two input streams, so both `sIn1' and `sIn2' map to `none'.

@@

Finally, note that in the result states @F0@ through @F3@ the state of the input streams transition from `none', to `pending' then to `have'. The `none' state means that we have not yet pulled a value from the associated stream. The `pending' state means we have pulled a value into the stream buffer variable (@b1@ in this case). The `have' state means that we have copied the pulled value from the stream buffer variable into the local variable used by each machine. In Fig.~\ref{fig:Fusion:Pulls},  `sIn1' is set to `have' for the first machine in @F2@ after we have set `v = b1', while `sIn1' is set to `have' for the second machine in @F3@ after we have set `x1 = b1'.

\begin{figure*}
\includegraphics[scale=1.1]{figures/fuse-pull-pull.pdf}
\caption{Fusing pull instructions}
\label{fig:Fusion:Pulls}
\end{figure*}

% -----------------------------------------------------------------------------
\subsubsection{Fusing Cases}
\begin{figure*}
\includegraphics[scale=1.1]{figures/fuse-case-pull.pdf}
\caption{Fusing case instructions}
\label{fig:Fusion:Case}
\end{figure*}

Once the result machine has arrived at the joint state @F3@, this is equivalent to the two source machines arriving in states @A1@ and @B1@ respectively. The left of Fig.~\ref{fig:Fusion:Case} shows the next few transitions of the source machines. From state @A1@, the @group@ machine needs to perform a @case@ branch to determine whether to push the current value it has from its input stream @sIn1@ to output stream @sOut1@, or to just pull the next value from its input. From state @B1@, the @merge@ machine needs to pull a value from its second input stream @sIn2@. In the result machine, @F3@ performs the case analysis from @A1@, moving to either @A2@ or @A3@, corresponding to @F4@ and @F5@ respectively. From state @F4@, the push at @A2@ is executed and moves to @A3@, corresponding to @F5@.

Finally, at @F5@ the @merge@ machine pulls from @sIn2@, moving from @B1@ to @C0@.
Because the stream @sIn2@ is only pulled from by the @merge@ machine, no coordination is required between the @merge@ and @group@ machines for this pull.

Note that we could construct the fused result machine in several ways. One option is to perform the case branch first and then pull from @sIn2@, another is to pull from @sIn2@ first and then perform the branch. By construction, the predicate used in the branch refers only to variables local to the @group@ machine, and the pull instruction from @B1@ stores its result in a variable local to the @merge@ machine. As the set of variables does not overlap, either ordering is correct. For this example we choose to perform the branch first, though will discuss the ramifications of this choice further in \S\ref{s:FusionOrder}.

@@

Fig.~\ref{fig:Process:Fused} shows the final result of fusing @group@ and @merge@ together. There are similar rules for handling the other combinations of instructions, but we defer the details to \S\ref{s:Fusion}. The result process has two input streams, @sIn1@ and @sIn2@ and two output streams: @sOut1@ which is @group@'s output stream, and @sOut2@ which is @merge@'s output stream.

@@

@@

\newcommand\tctt[2]{\textcolor{#1}{\tt{#2}}}
\begin{alltt}
process
\string{ ins:    \string{ \tctt{sharec}{sIn1},  \tctt{mergec}{sIn2} \string}
, outs:   \string{ \tctt{groupc}{sOut1}, \tctt{mergec}{sOut2} \string}
, heap:   \string{ \tctt{groupc}{f = T, l = 0, v = 0}, \tctt{mergec}{x1 = 0, x2 = 0}, \tctt{sharec}{b1 = 0} \string}
, label:  \tctt{sharec}{F0}
\end{alltt}

@@

\newcommand\icase[7]{
 \tt{#1} \> \tt{= #2} \> \tt{#3} \> \tt{[ #4 ]} \> \tt{#5} \> \tt{[ #6 ]} \> \tiny #1 \> \tiny = #7
}
\newcommand\instr[5]{
 \tt{#1}\>\tt{= #2} \> \tt{#3} \> \tt{[ #4 ]} \> \> \> \tiny #1 \> \tiny = #5
}

@@

\\[0pt \color{black}]
@} }@
\end{tabbing}

\caption{Fusion of \textcolor{groupc}{group} and \textcolor{mergec}{merge}, along with \textcolor{sharec}{shared} instructions}
\label{fig:Process:Fused}
\end{figure*}

>>>>

% -----------------------------------------------------------------------------
\section{Processes and Machines}
\label{s:Processes}

@@

The definition of the @group@ combinator which detects groups of successive identical elements in the input stream is given in Figure~\ref{fig:Process:Group}. The process emits the first value pulled from the stream and every value that is different from the last one that was pulled. For example, when executed on the input stream @[1,2,2,3]@, the process will produce the output @[1,2,3]@. We include the concrete representation and a diagram of the process viewed as a state machine.

@@

The body of the combinator is a record that defines the process. The @ins@ field defines the set of input streams and the @outs@ field the set of output streams. The @heap@ field gives the initial values of each of the local variables. The @instrs@ field contains a set of labeled instructions that define the program, while the @label@ field gives the label of the initial instruction. In this form the output stream is defined via a parameter, rather than being the result of the combinator, as in the representation of @uniquesUnion@ from \S\ref{s:Introduction}.

@@

% AR: Clumsy, but added "at L2 executes:" to force @(push...)@ onto its own line. Otherwise it goes off the right.
When the predicate is true, the instruction at label @A2@ executes:
@(push sOut1 v A3 [ l = v, f = F ])@
which pushes the value @v@ to the output stream @sOut1@ and proceeds to the instruction at label @A3@, once the variable @l@ is set to @v@ and @f@ to @F@ (False).

Finally, the instruction @(drop sIn1 A0 [])@ signals that the current element that was pulled from stream @sIn1@ is no longer required, and goes back to the first the instruction at @A0@.

% TODO: Talk about what drop is really for later.
% This @drop@ instruction is used to coordinate concurrent processes when performing fusion. The next element of a stream may only be pulled after all consumers of that stream have pulled and then and dropped the current element.

@@

The definition of the @merge@ combinator, which merges two input streams, is given in Figure~\ref{fig:Process:Merge}. The combinator binds the two input streams to @sIn1@ and @sIn2@, while the output stream is @sOut2@. The two heap variables @x1@ and @x2@ are used to store the last values read from each input stream. The process starts by pulling from each of the input streams. It then compares the two pulled values, and pushes the smaller of the values to the output stream. The process then drops the stream which yielded the the smaller value, then pulls from the same stream so that it can perform the comparison again.

% TODO: Shift this later.
% As the @merge@ process merges infinite streams, if we execute it with a finite input prefix, it will arrive at an intermediate state that may not yet have pushed all available output. For example, if we execute the process with the input streams @[1,4]@ and @[2,3,100]@ then the values @[1,2,3,4]@ will be pushed to the output. After pushing the last value @4@, the process will block at instruction @E2@, waiting for the next value to become available from @sIn2@. We discuss how to handle finite streams later in ~\S\ref{s:Finite}.

@@

Our fusion algorithm takes two processes and produces a new one that computes the output of both. For example, suppose we need a single process that produces the output of the first two lines of our @uniquesUnion@ example back in \S\ref{s:Introduction}. The result will be a process that computes the result of both @group@ and @merge@ as if they were executed concurrently, where the first input stream of the @merge@ process is the same as the input stream of the @group@ process. When describing the fusion algorithm we will instantiate the parameters of each combinator with arguments of the same names.

@@

The algorithm proceeds by considering pairs of states: one in each of the source process state machines to be fused. Both the @group@ machine and the @merge@ machine pull from the same stream as their initial instruction, so we have the situation shown in Figure~\ref{fig:Fusion:Pulls}. The @group@ machine needs to transition from label @A0@ to label @A1@, and the @merge@ machine from @B0@ to @B1@. In the result machine we produce three new instructions that transition between four joint result states, @F0@ to @F3@.
Each of the joint result states represents a combination of two source states, one from each of the source machines. For example, the first result state @F0@ represents a combination of the @group@ machine being in its initial state @A0@ and the @merge@ machine being in its own initial state @B0@.

We also associate each of the joint result states with whether or not each source machine has already pulled a value from each of its input streams. For the @F0@ case shown in Figure~\ref{fig:Fusion:Pulls} we have ((A0, \{sIn1 = none\}), (B0, \{sIn1 = none, sIn2 = none\})). The result state @F0@ represents a combination of the two source states @A0@ and @B0@. As both @A0@ and @B0@ are the initial states of their respective machines, those machines have not yet pulled any values from their two input streams, so both `sIn1' and `sIn2' map to `none'.

@@

Finally, note that in the result states @F0@ through @F3@ the state of the input streams transition from `none', to `pending' then to `have'. The `none' state means that we have not yet pulled a value from the associated stream. The `pending' state means we have pulled a value into the stream buffer variable (@b1@ in this case). The `have' state means that we have copied the pulled value from the stream buffer variable into the local variable used by each machine.

% BL: Dropped this to fix pagination.
% In Figure~\ref{fig:Fusion:Pulls},  `sIn1' is set to `have' for the first machine in @F2@ after we have set `v = b1', while `sIn1' is set to `have' for the second machine in @F3@ after we have set `x1 = b1'.

% -----------------------------------------------------------------------------
\subsubsection{Fusing Cases}
Once the result machine has arrived at the joint state @F3@, this is equivalent to the two source machines arriving in states @A1@ and @B1@ respectively. The left of Figure~\ref{fig:Fusion:Case} shows the next few transitions of the source machines. From state @A1@, the @group@ machine needs to perform a @case@ branch to determine whether to push the current value it has from its input stream @sIn1@ to output stream @sOut1@, or to just pull the next value from its input. From state @B1@, the @merge@ machine needs to pull a value from its second input stream @sIn2@. In the result machine, @F3@ performs the case analysis from @A1@, moving to either @A2@ or @A3@, corresponding to @F4@ and @F5@ respectively. From state @F4@, the push at @A2@ is executed and moves to @A3@, corresponding to @F5@. Finally, at @F5@ the @merge@ machine pulls from @sIn2@, moving from @B1@ to @C0@. Because the stream @sIn2@ is only pulled from by the @merge@ machine, no coordination is required between the @merge@ and @group@ machines for this pull.

@@

% TODO: Blend this in.
Note that we could construct the fused result machine in several ways. One option is to perform the case branch first and then pull from @sIn2@, another is to pull from @sIn2@ first and then perform the branch. By construction, the predicate used in the branch refers only to variables local to the @group@ machine, and the pull instruction from @B1@ stores its result in a variable local to the @merge@ machine. As the set of variables does not overlap, either ordering is correct. For this example we choose to perform the branch first, though will discuss the ramifications of this choice further in \S\ref{s:FusionOrder}.

Figure~\ref{fig:Process:Fused} shows the final result of fusing @group@ and @merge@ together. There are similar rules for handling the other combinations of instructions, but we defer the details to \S\ref{s:Fusion}. The result process has two input streams, @sIn1@ and @sIn2@ and two output streams: @sOut1@ which is @group@'s output stream, and @sOut2@ which is @merge@'s output stream.

@@

% -----------------------------------------------------------------------------
\begin{figure*}
\includegraphics[scale=1.1]{figures/fuse-pull-pull.pdf}

\vspace{2em}

\includegraphics[scale=1.1]{figures/fuse-case-pull.pdf}
\caption{Fusing pull (top) and case (bottom) instructions}
\label{fig:Fusion:Pulls}
\label{fig:Fusion:Case}
\end{figure*}

% BL: Merged these figures togather to save space.
% \begin{figure*}
% \caption{Fusing case instructions}
% \label{fig:Fusion:Case}
% \end{figure*}

@@

@@

\newcommand\icase[7]{
 \tt{#1} \> \tt{= #2} \> \tt{#3} \> \tt{[ #4 ]} \> \tt{#5} \> \tt{[ #6 ]} \> \tiny #1 \> \tiny = #7
}

\newcommand\instr[5]{
 \tt{#1}\>\tt{= #2} \> \tt{#3} \> \tt{[ #4 ]} \> \> \> \tiny #1 \> \tiny = #5
}

\newcommand\tctt[2]{\textcolor{#1}{\tt{#2}}}

\hspace{5em}

\begin{adjustbox}{minipage=0.8\textwidth,margin=0pt \smallskipamount,center}
\begin{alltt}
process
\string{ ins:    \string{ \tctt{sharec}{sIn1},  \tctt{mergec}{sIn2} \string}
, outs:   \string{ \tctt{groupc}{sOut1}, \tctt{mergec}{sOut2} \string}
, heap:   \string{ \tctt{groupc}{f = T, l = 0, v = 0}, \tctt{mergec}{x1 = 0, x2 = 0}, \tctt{sharec}{b1 = 0} \string}
, label:  \tctt{sharec}{F0}
\end{alltt}

@@

\\[0pt \color{black}]
@} }@
\end{tabbing}
\end{adjustbox}
\caption{Fusion of \textcolor{groupc}{group} and \textcolor{mergec}{merge}, along with \textcolor{sharec}{shared} instructions}
\label{fig:Process:Fused}
\end{figure*}
% -----------------------------------------------------------------------------
